Joint estimation of phylogenies and ancestral states is not consistent

### 2017-12-31 appendix.tex
Finished pass through appendix.

\subsection*{Site split formulation}

Since we have a finite character alphabet, for a given column $i$ there are a finite number of possible assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$; this results in a simplification of likelihood calculation.
Take the tip labels of $\tau$ to be $\{1,\ldots,\nSiteRows\}$.
For likelihood calculation under the binary symmetric model, we describe a given $\alignmentColumn_i$ as a subset of indices $\siteSplit\subseteq\siteSplitSet:=\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
We start by including those tip labels in $\siteSplit$ that are assigned a 1 in the $\alignmentColumn_i$.
Then, because the probability of observing a particular collection of binary characters is equivalent to the probability of its complement under the binary symmetric model, we take the complement if needed to exclude label $\nSiteRows$ from $\siteSplit$.

@@

For the $i$th factor of \eqref{eq:full_likelihood},
$$
P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = P(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) \cdot P(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t).
$$
As a consequence of assuming a binary symmetric model, taking complements yields
%EM continuing to nitpick, is there anything that is just s_j? Seems funny to only have the tilde s_j form.
%das: I think s_j used to be the site pattern and then \tilde{s}_j became the split? It's still the case for h (no tilde is ancestral state pattern, tilde is ancestral state split). I changed to \tilde{y} to be consistent with this, but mabye we can think of something better?
\begin{align*}
    2\cdot P(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) &= P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \\
                                                                 &= P(\siteSplitRV=\siteSplit_j \mid \tau, t)
\end{align*}
for some $j$ and
\begin{align*}
    P(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t).
\end{align*}
Given $(\tau, t)$, there exists an ordered list of sets $\fullAncestralSplitPartitions(\tau, t)=(\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t))$ such that any element $\xi_j$ of the $j$th component $\ancestralSplitPartition_j(\tau, t)$ satisfies
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= P(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}
In other words, for the $j$th site split, $\ancestralSplitPartition_j(\tau, t)\subset\mathcal{P}(\ancestralSplitSet)$ is the set of most likely ancestral splits for that particular site split, topology and set of branch lengths, and $\xi_j$ is one of possibly many equiprobable ancestral state splits in $\ancestralSplitPartition_j(\tau, t)$.
For each $\alignmentColumn_i$, $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective, and from this we have
\begin{align*}
\max_{\ancestralStateColumn_i} P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) &= P(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}

@@

Let $\xi_j$ be such a choice for each $1 \leq j \leq q$ in the following.
Then, the likelihood in \eqref{eq:profile_likelihood} written as a product over site patterns as opposed to sites is
\begin{align}
L_\nCols'(\tau, t; \fullAlignment) &= \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \nonumber \\
                             &\propto \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \max_{\ancestralStateColumn_i} P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
                             &= \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ where the site split $\siteSplit_j$ or its complement appears.

Let
$$
%EM It seems a little strange that eta appears here as an argument when it didn't appear in eq:site_pattern_likelihood. In fact it's not really an argument, right? Everything is determined by t and tau it seems to me.
%das: We added it to clarify that the likelihood now depends on \xi, which is an element of \eta. It seems dirty to ignore it. At the risk of adding more notation, we could add \hat{H} to the original likelihood? Or just axe the etas...
%EM I think I must be missing something:
%EM The likelihood doesn't belong on the specific choice of \xi. Besides, \xi already appears in eq:site_pattern_likelihood.
%EM Eta is a deterministic function of t and tau, so isn't having it redundant?
%das: The likelihood doesn't depend on the specific choice of \xi, but it does depend on \eta. I don't know, I guess redundancy isn't always the worst thing? I just see it as a tradeoff between how explicit we want to be and how clean. Like, in the first line of eq:site_pattern_likelihood we could have as an argument \hat{H} instead of suppressing it, even though it's totally determined by the inputs as-is. Anyway, that was my justification for including \eta. I'm fine with removing it though if you think it's weird.
%das: I changed all this.
L_\nCols''(\tau, t; \fullAlignment) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
$$
be the final product in \eqref{eq:site_pattern_likelihood}.
Assume $\nCols$ observations are generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{equation*}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) +
            \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\end{split}
\end{equation*}
so that, in the $\nCols\rightarrow\infty$ limit,
\begin{equation}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
&\qquad \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{split}
\end{equation}
Define the divergence quantity
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\siteSplitRV=\siteSplit_j \mid \tau, t)
$$
and the partial log-likelihood
$$
\tilde{\ell}_{\tau^*,t^*}(\tau, t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, t)
$$
so that \eqref{eq:site_pattern_profile_likelihood_mean} becomes
\begin{equation}
    \label{eq:log_likelihood_simplified}
    \ell_{\tau^*,t^*}(\tau, t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \tilde{\ell}_{\tau^*,t^*}(\tau, t).
\end{equation}

@@

We follow Section 8.6 of \cite{Semple2003-em} to state the Hadamard representation of site split generating probabilities.
For each edge $e$ define the edge fidelity for that edge as
$$
\theta(e) = 1-2p(e).
$$
For an even-sized subset of $Y\subseteq\mathcal{S}$, we define the path set $P(Y)$ as the set of edges in the path connecting both elements of $Y$.
For $n$ taxa, the probability of observing site split $A\in\mathcal{P}(\siteSplitSet)$ is
\begin{equation}
\label{eq:hadamard_probability}
p_A = \frac{1}{2^{n-1}} \ \sum_{Y \subseteq \mathcal{S} : |Y| \equiv 0 (\mathrm{mod} \ 2)} \ \left[(-1)^{|Y \cap A|} \ \prod_{e\in P(Y)} \ \theta(e) \right].
\end{equation}
By convention, we set $P(\emptyset)=\emptyset$ and $\prod_{e\in\emptyset} \ \theta(e) = 1$.
Table~\ref{tab:sitepatprob} contains calculations of site pattern frequencies for our two topologies (Fig.~\ref{fig:farris-fels-top}).

\begin{table}[ht]
\centering
\begin{tabular}{|l|l|l|}
    \hline
$\siteSplit_j$   &$P(\siteSplitRV=\siteSplit_j \mid \tau_1,t)$&$P(\siteSplitRV=\siteSplit_j \mid \tau_2,t)$\\
    \hline
    $\emptyset$    &$1+x^2+y^2+4xyw+x^2y^2$&$1+2xy+2xyw+x^2w+y^2w+x^2y^2$\\
    $\{1\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
    $\{2\}$        &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
    $\{3\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
    $\{1,2\}$      &$1-x^2-y^2+x^2y^2$&$1+2xy-2xyw-x^2w-y^2w+x^2y^2$\\
    $\{1,3\}$      &$1+x^2+y^2-4xyw+x^2y^2$&$1-2xy-2xyw+x^2w+y^2w+x^2y^2$\\
    $\{2,3\}$      &$1-x^2-y^2+x^2y^2$&$1-2xy+2xyw-x^2w-y^2w+x^2y^2$\\
    $\{1,2,3\}$    &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
    \hline
\end{tabular}
\caption{Site pattern probabilities $\siteSplit_j$ on the Farris tree $\tau_1$ and the Felsenstein tree $\tau_2$ obtained using the Hadamard transform.
All values multiplied by $1/8$.}
\label{tab:sitepatprob}
\end{table}

@@

\begin{table}
\centering
\begin{tabular}{|l|ll|}
\multicolumn{3}{c}{$P(\ancestralStateColumn=\ancestralSplit_k\mid \alignmentColumn=\siteSplit_j,\tau_1,t)$}\\
\hline
& \multicolumn{2}{|c|}{$\ancestralSplit_k$}\\
    \hline
    $\siteSplit_j$    &$\emptyset$                                &$\{2\}$  \\
    \hline
     $\emptyset$   &$(1+x)^2   (1+w)(1+y)^2$          &$(1+x)^2   (1-w)(1-y)^2$\\
     $\{1\}$       &$(1+x)(1-x)(1+w)(1+y)^2$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
     $\{2\}$       &$(1+x)^2   (1+w)(1+y)(1-y)$       &$(1+x)^2   (1-w)(1+y)(1-y)$\\
     $\{3\}$       &$(1+x)(1-x)(1+w)(1+y)^2$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
     $\{1,2\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$       &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
     $\{1,3\}$     &$(1-x)^2   (1+w)(1+y)^2$          &$(1-x)^2   (1-w)(1-y)^2$\\
     $\{2,3\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$       &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
     $\{1,2,3\}$   &$(1-x)^2   (1+w)(1+y)(1-y)$       &$(1-x)^2   (1-w)(1+y)(1-y)$\\
    \hline
    \hline
    &$\{1\}$                             &$\{1,2\}$  \\
    \hline
     $\emptyset$   &$(1-x)^2   (1-w)(1+y)^2$     &$(1-x)^2   (1+w)(1-y)^2$\\
     $\{1\}$       &$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
     $\{2\}$       &$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
     $\{3\}$       &$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
     $\{1,2\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
     $\{1,3\}$     &$(1+x)^2   (1-w)(1+y)^2$     &$(1+x)^2   (1+w)(1-y)^2$\\
     $\{2,3\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
     $\{1,2,3\}$   &$(1+x)^2   (1-w)(1+y)(1-y)$  &$(1+x)^2   (1+w)(1+y)(1-y)$\\
    \hline
\end{tabular}
\caption{Likelihood calculations for all site patterns $\siteSplit_j$ and internal states $\ancestralSplit_k$ of the Farris tree.
All values multiplied by $1/32$.}
\label{tab:farris_likelihoods}
\end{table}

>>>>

\subsection*{Site split formulation}
We begin by introducing ``site splits,'' that simply formalize the notion that a given site pattern is equally probable to its complement under the binary symmetric model.
This is a standard step in the description of the Hadamard transform (Section 8.6 of \cite{Semple2003-em}), although we have to be a little careful here due to the extra bookkeeping of ancestral states.

Since we have a finite character alphabet, for a given column $i$ there are a finite number of possible assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$; this results in a simplification of likelihood calculation.
Take the tip labels of $\tau$ to be $\{1,\ldots,\nSiteRows\}$.
For likelihood calculation under the binary symmetric model, we describe a given $\alignmentColumn_i$ as a subset of indices $\siteSplit\subseteq\siteSplitSet:=\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
We define the site split $\siteSplit$ for a $\alignmentColumn_i$ as simply $\alignmentColumn_i$ if the label $\nSiteRows$ is not in $\alignmentColumn_i$, and as its complement otherwise.
Taking such a complement simplifies but does not change the result of likelihood computation because the probability of observing a particular collection of binary characters is equivalent to the probability of its complement under the binary symmetric model.

@@

For the $i$th factor of \eqref{eq:full_likelihood},
$$
P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = P(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) \cdot P(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t).
$$
As a consequence of assuming a binary symmetric model, taking complements yields
\begin{align*}
    2\cdot P(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) &= P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \\
                                                                 &= P(\siteSplitRV=\siteSplit_j \mid \tau, t)
\end{align*}
for some $j$ and
\begin{align*}
    P(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t).
\end{align*}
Given $(\tau, t)$, there exists an ordered list of sets $\fullAncestralSplitPartitions(\tau, t)=(\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t))$ such that any element $\xi_j$ of the $j$th component $\ancestralSplitPartition_j(\tau, t)$ satisfies
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= P(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}
In other words, for the $j$th site split, $\ancestralSplitPartition_j(\tau, t)\subset\mathcal{P}(\ancestralSplitSet)$ is the set of most likely ancestral splits for that particular site split, topology and set of branch lengths, and $\xi_j$ is one of possibly many equiprobable ancestral state splits in $\ancestralSplitPartition_j(\tau, t)$.
For each $\alignmentColumn_i$, $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective, and from this we have
\begin{align*}
\max_{\ancestralStateColumn_i} P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) &= P(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}

@@

Let $\xi_j$ be such a choice for each $1 \leq j \leq q$ in the following.
Then, the likelihood in \eqref{eq:profile_likelihood} written as a product over site patterns as opposed to sites is
\begin{align}
L_\nCols'(\tau, t; \fullAlignment) &= \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \nonumber \\
                             &\propto \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \max_{\ancestralStateColumn_i} P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
                             &= \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ that project to site split $\siteSplit_j$.

Let
$$
L_\nCols''(\tau, t; \fullAlignment) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
$$
be the final product in \eqref{eq:site_pattern_likelihood}.
Assume $\nCols$ observations are generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{equation*}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) +
            \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\end{split}
\end{equation*}
so that, in the $\nCols\rightarrow\infty$ limit,
\begin{equation}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
&\qquad \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{split}
\end{equation}
Define the divergence quantity
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\siteSplitRV=\siteSplit_j \mid \tau, t)
$$
and the partial log-likelihood
$$
\tilde{\ell}_{\tau^*,t^*}(\tau, t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, t)
$$
so that \eqref{eq:site_pattern_profile_likelihood_mean} becomes
\begin{equation}
    \label{eq:log_likelihood_simplified}
    \ell_{\tau^*,t^*}(\tau, t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \tilde{\ell}_{\tau^*,t^*}(\tau, t).
\end{equation}

@@

We now state the Hadamard representation of site split generating probabilities, following Section 8.6 of \cite{Semple2003-em}.
For each edge $e$ define the edge \emph{fidelity} for that edge as
$$
\theta(e) = 1-2p(e).
$$
For an even-sized subset of $Y\subseteq\mathcal{S}$, we define the path set $P(Y)$ as the set of edges in the path connecting both elements of $Y$.
For $n$ taxa, the probability of observing site split $A\in\mathcal{P}(\siteSplitSet)$ is
\begin{equation}
\label{eq:hadamard_probability}
p_A = \frac{1}{2^{n-1}} \ \sum_{Y \subseteq \mathcal{S} : |Y| \equiv 0 (\mathrm{mod} \ 2)} \ \left[(-1)^{|Y \cap A|} \ \prod_{e\in P(Y)} \ \theta(e) \right].
\end{equation}
By convention, we set $P(\emptyset)=\emptyset$ and $\prod_{e\in\emptyset} \ \theta(e) = 1$.
Table~\ref{tab:sitepatprob} contains calculations of site pattern frequencies for our two topologies (Fig.~\ref{fig:farris-fels-top}).

\begin{table}[ht]
\centering
\begin{tabular}{|l|l|l|}
    \hline
$\siteSplit_j$   &$P(\siteSplitRV=\siteSplit_j \mid \tau_1,t)$&$P(\siteSplitRV=\siteSplit_j \mid \tau_2,t)$\\
    \hline
    $\emptyset$    &$1+x^2+y^2+4xyw+x^2y^2$&$1+2xy+2xyw+x^2w+y^2w+x^2y^2$\\
    $\{1\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
    $\{2\}$        &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
    $\{3\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
    $\{1,2\}$      &$1-x^2-y^2+x^2y^2$&$1+2xy-2xyw-x^2w-y^2w+x^2y^2$\\
    $\{1,3\}$      &$1+x^2+y^2-4xyw+x^2y^2$&$1-2xy-2xyw+x^2w+y^2w+x^2y^2$\\
    $\{2,3\}$      &$1-x^2-y^2+x^2y^2$&$1-2xy+2xyw-x^2w-y^2w+x^2y^2$\\
    $\{1,2,3\}$    &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
    \hline
\end{tabular}
%EM When I came back to this after the break I found it confusing that we refer to these as p_A below (eg in proof of Theorem 1) but that notation doesn't appear in this table. Can we use it here?
\caption{Site pattern probabilities $\siteSplit_j$ on the Farris tree $\tau_1$ and the Felsenstein tree $\tau_2$ obtained using the Hadamard transform.
All values multiplied by $1/8$.}
\label{tab:sitepatprob}
\end{table}

@@

\begin{table}
\centering
\begin{tabular}{|l|ll|}
\multicolumn{3}{c}{$P(\ancestralStateColumn=\ancestralSplit_k\mid \alignmentColumn=\siteSplit_j,\tau_1,t)$}\\
\hline
& \multicolumn{2}{|c|}{$\ancestralSplit_k$}\\
    \hline
    $\siteSplit_j$    &$\emptyset$                                &$\{2\}$  \\
    \hline
     $\emptyset$   &$(1+x)^2   (1+w)(1+y)^2$          &$(1+x)^2   (1-w)(1-y)^2$\\
     $\{1\}$       &$(1+x)(1-x)(1+w)(1+y)^2$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
     $\{2\}$       &$(1+x)^2   (1+w)(1+y)(1-y)$       &$(1+x)^2   (1-w)(1+y)(1-y)$\\
     $\{3\}$       &$(1+x)(1-x)(1+w)(1+y)^2$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
     $\{1,2\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$       &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
     $\{1,3\}$     &$(1-x)^2   (1+w)(1+y)^2$          &$(1-x)^2   (1-w)(1-y)^2$\\
     $\{2,3\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$       &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
     $\{1,2,3\}$   &$(1-x)^2   (1+w)(1+y)(1-y)$       &$(1-x)^2   (1-w)(1+y)(1-y)$\\
    \hline
    \hline
    &$\{1\}$                             &$\{1,2\}$  \\
    \hline
     $\emptyset$   &$(1-x)^2   (1-w)(1+y)^2$     &$(1-x)^2   (1+w)(1-y)^2$\\
     $\{1\}$       &$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
     $\{2\}$       &$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
     $\{3\}$       &$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
     $\{1,2\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
     $\{1,3\}$     &$(1+x)^2   (1-w)(1+y)^2$     &$(1+x)^2   (1+w)(1-y)^2$\\
     $\{2,3\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
     $\{1,2,3\}$   &$(1+x)^2   (1-w)(1+y)(1-y)$  &$(1+x)^2   (1+w)(1+y)(1-y)$\\
    \hline
\end{tabular}
\caption{Likelihood calculations for all site patterns $\siteSplit_j$ and internal states $\ancestralSplit_k$ of the Farris tree $\tau_1$.
All values multiplied by $1/32$.}
\label{tab:farris_likelihoods}
\end{table}

\begin{table}
\centering
\begin{tabular}{|l|ll|}
\multicolumn{3}{c}{$P(\ancestralStateColumn=\ancestralSplit_k\mid \alignmentColumn=\siteSplit_j,\tau_2,t)$}\\
\hline
& \multicolumn{2}{|c|}{$\ancestralSplit_k$}\\
    \hline
    $\siteSplit_j$    &$\emptyset$                                &$\{2\}$  \\
    \hline
     $\emptyset$   &$(1+x)^2   (1+w)(1+y)^2$           &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
     $\{1\}$       &$(1+x)(1-x)(1+w)(1+y)^2$           &$(1-x)^2   (1-w)(1+y)(1-y)$\\
     $\{2\}$       &$(1+x)^2   (1+w)(1+y)(1-y)$        &$(1+x)(1-x)(1-w)(1-y)^2$\\
     $\{3\}$       &$(1+x)(1-x)(1+w)(1+y)^2$           &$(1+x)^2   (1-w)(1+y)(1-y)$\\
     $\{1,2\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$        &$(1-x)^2   (1-w)(1-y)^2$\\
     $\{1,3\}$     &$(1-x)^2   (1+w)(1+y)^2$           &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
     $\{2,3\}$     &$(1+x)(1-x)(1+w)(1+y)(1-y)$        &$(1+x)^2   (1-w)(1-y)^2$\\
     $\{1,2,3\}$   &$(1-x)^2   (1+w)(1+y)(1-y)$        &$(1+x)(1-x)(1-w)(1-y)^2$\\
    \hline
    \hline
    &$\{1\}$                             &$\{1,2\}$  \\
    \hline
     $\emptyset$   &$(1+x)(1-x)(1-w)(1+y)(1-y)$        &$(1-x)^2   (1+w)(1-y)^2$\\
     $\{1\}$       &$(1+x)^2   (1-w)(1+y)(1-y)$        &$(1+x)(1-x)(1+w)(1-y)^2$\\
     $\{2\}$       &$(1+x)(1-x)(1-w)(1+y)^2$           &$(1-x)^2   (1+w)(1+y)(1-y)$\\
     $\{3\}$       &$(1-x)^2   (1-w)(1+y)(1-y)$        &$(1+x)(1-x)(1+w)(1-y)^2$\\
     $\{1,2\}$     &$(1+x)^2   (1-w)(1+y)^2$           &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
     $\{1,3\}$     &$(1+x)(1-x)(1-w)(1+y)(1-y)$        &$(1+x)^2   (1+w)(1-y)^2$\\
     $\{2,3\}$     &$(1-x)^2   (1-w)(1+y)^2$           &$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
     $\{1,2,3\}$   &$(1+x)(1-x)(1-w)(1+y)^2$           &$(1+x)^2   (1+w)(1+y)(1-y)$\\
\hline
\end{tabular}
\caption{Likelihood calculations for all site patterns $\siteSplit_j$ and internal states $\ancestralSplit_k$ of the Felsenstein tree $\tau_2$.
All values multiplied by $1/32$.}
\label{tab:fels_likelihoods}
\end{table}
